Sexual selection, environmental variation, range size and speciation in passerine

### 2018-10-11 Manuscript.Rmd
Luke MS edits 1

---
title: Sexual selection, environmental variation, range size and speciation in passerine
  birds
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§} and Iliana Medina Guzmán\textsuperscript{§}
- ~
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  html_document: default
  word_document:
    pandoc_args: --smart
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual Selection and Speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{hyperref}\hypersetup{colorlinks=true, linkcolor=blue, filecolor=blue, urlcolor=blue}
  \raggedbottom
csl: References/AmNat.csl
abstract: 'Sexual selection may influence patterns of diversification; yet controversy remains on whether sexual selection leads to a net increase or decrease in speciation rate, or even elevates extinction. Sexual selection is predicted to increase speciation rate as it can facilitate adaptation to new environments, and work alongside divergent natural selection to swiftly promote reproductive isolation. However, sexual selection frequently favours investment in costly or harmful traits that may reduce adaptation and affect diversification. Here, we investigate the association of sexual selection with speciation and extinction in passerine birds. We use two measures of sexual selection: sexual dichromatism (n = 5,812 species) and a multivariate measure of male-biased sexual selection (sexual dimorphism, polygyny and paternal care; n = 2,465), three distinct measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$) and one measure of extinction ($\mu_{BAMM}$). Importantly, we test whether the effects of sexual selection on speciation are dependent upon ecological pressures such as temperature variability across time and space, primary productivity and range size. We find that male-biased sexual selection but not sexual dichromatism predicts speciation. Additionally, we find that birds with smaller ranges have higher speciation rates; although the direction of causality is unclear.'
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Selection pressures are the main engine of speciation. As populations diverge they accumulate genetic and phenotypic changes that ultimately erect barriers against gene flow [@Presgraves_2010; *but see* @Shaw_2011]. In this view, populations will speciate and adapt more readily to new and different environments if selection is strong and/or persistant. Hence, the rate at which speciation occurs is likely dependent on both the populations response to selection (adaptation) and the potential for divergent phenotypic traits to evolve [*e.g.* morphology or colour @Hugall_2012; @Rabosky_morph_2013]. Sexual selection is an evolutionary force likely to affect fitness components of a population [@Cally_in_review], such as reproductive success or viability, as well as the evolution of ornaments involved in mating displays or weapons used in male-male competition [@McCullough_2016]. Sexual selection is therefore hypothesised to be a key driver of diversification through shifting the rates of adaptation, speciation and extinction.

Sexual selection is applied by one sex on the other --- more often by females on males [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008] and the strength of sexual selection can increase or decrease the average fitness of a population; often termed population fitness [*reviewed in* @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. Specifically, in line with the 'good genes' perspective, sexual selection may purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial mutations in a population [@Whitlock_2000]. Thus, sexual selection is thought to be a key process in mitigating extinction risk and facilitating adaptation. Sexual selection can also increase the fitness of a population when the genotypes under sexual selection (*e.g.* those that increase male mating/fertilisation success) are genetically positively correlated with traits affecting fitness (*e.g.* female fecundity or survival). However, this positive correlation is not ubiquitous; indeed sexual selection frequently favours traits having negative pleiotropic effects on female fitness [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006]. For instance, in species with high levels of sexual selection, mating success becomes more variable and uncertain; thus individuals must invest heavily in costly weaponry and sexual signals to increase their chances of successfully reproducing. Such high investment, may trade off against strategies that would improve offspring fitness (*e.g.* parental care). This is a costly consequence of sexual selection that affects persistence, adaptation and extinction [@Kokko_Brooks_2003].

@@

![Model estimates for (**a**) PGLS analyses using sexual dichromatism (*n* = 5,812) and (**b**) PGLS analyses using measures of male-biased sexual selection (*n* = 2,465). Both datasets were used for analyses with three measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$) and one measure of extinction ($\mu_{BAMM}$) as response variables. The numerical values for the model estimates using the MCC tree and HPD intervals of estimates from 100 random trees can be found within the ESM. Density curves are based on estimates from 100 trees and the circle below with error bars is the estimate and 95 % CIs from the MCC tree. For this figure we removed outliers from estimates coming from the 100 random trees for BAMM models in order to interpret the MCC 95 % CIs. \label{model_results}](Figures/ModelEstimates.pdf)

>>>>

---
title: Sexual selection, environmental variation, range size and speciation in passerine
  birds
author:
- Justin G. Cally*\textsuperscript{§}, Devi Stuart-Fox\textsuperscript{§}, Luke Holman\textsuperscript{§} and Iliana Medina Guzmán\textsuperscript{§}
- ~
- '*\textit{justin.g.cally@gmail.com}'
bibliography: References/SS_Extinction.bib
output:
  pdf_document: default
  html_document: default
  word_document:
    pandoc_args: --smart
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Sexual Selection and Speciation} \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
  \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption}
  \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
  \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip}
  \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape}
  \usepackage{hyperref}\hypersetup{colorlinks=true, linkcolor=blue, filecolor=blue, urlcolor=blue}
  \raggedbottom
csl: References/AmNat.csl
abstract: 'Many theories related to sexual selection predict that it shapes phylogenetic diversity by altering speciation or extinction rates. The predicted effects on diversity are often contradictory; for example, sexual selection might promote speciation by creating reproductive isolation, or hinder it by selecting for traits that facilitate gene flow. Similarly, sexual selection might precipitate extinction by favouring competitive phenotypes with poor survival, or stave it off by helping populations shed deleterious mutations. Here, we investigate the relationship between sexual selection and rates of speciation and extinction in passerine birds. We use two measures of sexual selection -- sexual dichromatism (n = 5,812 species) and a multivariate measure of male-biased sexual selection (sexual dimorphism, polygyny and paternal care; n = 2,465) -- as well as three distinct measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$), and one measure of extinction ($\mu_{BAMM}$). We also tested whether the effects of sexual selection is contingent on ecological variables such as spatiotemporal temperature variation, primary productivity, and range size. We found a positive correlation between speciation rate and male-biased sexual selection, as well as a negative correlation between range size and speciation rate. **ONE MORE SENTENCE HERE**'
---
\maketitle\unmarkedfntext{\textsuperscript{§}School of BioSciences, The University of Melbourne, Victoria, Australia.}

@@

Selection pressures are the main engine of speciation. As populations diverge they accumulate genetic and phenotypic changes that ultimately erect barriers against gene flow [@Presgraves_2010; *but see* @Shaw_2011]. In this view, populations will speciate and adapt more readily to new and different environments if selection is strong and/or persistant. Hence, the rate at which speciation occurs is likely dependent on both the population's response to selection (adaptation) and the potential for divergent phenotypic traits to evolve [*e.g.* morphology or colour @Hugall_2012; @Rabosky_morph_2013]. Sexual selection affects a diverse range of phenotypic traits important to population growth and persistance [@Cally_in_review], such as reproductive success or viability, as well as the evolution of ornaments involved in mating displays or weapons used in male-male competition [@McCullough_2016]. Sexual selection is therefore hypothesised to be a key driver of diversification, via its effects on adaptation, speciation and extinction.

Sexual selection results from competition for matings, fertilisations, or associated resources, and is typically stronger in males than females [@Fromhage_2016; @Janicke_2016; @Kokko_Jennions2008]. Sexual selection has both positive and negative consequences for 'population fitness', i.e. the average absolute fitness of individuals in the population [e.g. @Kokko_Brooks_2003; @Whitlock_Agrawal_2009; @Holman_Kokko_2013]. For example, sexual selection can elevate population fitness by helping to purge deleterious mutations [@Agrawal_2001; @Siller_2001] and fix beneficial ones [@Whitlock_2000], which should mitigate the risk of extinction due to mutation accumulation or a failure to adapte to changing environments. However, sexual selection does not always favour the evolution of traits that benefit the population...

this positive correlation is not ubiquitous; indeed sexual selection frequently favours traits having negative pleiotropic effects on female fitness [@Berger_2014; @Bonduriansky_Chenoweth_2009; @Pennell_2013; @Pischedda_2006]. For instance, in species with high levels of sexual selection, mating success becomes more variable and uncertain; thus individuals must invest heavily in costly weaponry and sexual signals to increase their chances of successfully reproducing. Such high investment, may trade off against strategies that would improve offspring fitness (*e.g.* parental care). This is a costly consequence of sexual selection that affects persistence, adaptation and extinction [@Kokko_Brooks_2003].

@@

![Model estimates for (**a**) PGLS analyses using sexual dichromatism (*n* = 5,812) and (**b**) PGLS analyses using measures of male-biased sexual selection (*n* = 2,465). Both datasets were used for analyses with three measures of speciation ($\lambda_{DR}$, $\lambda_{ND}$, $\lambda_{BAMM}$) and one measure of extinction ($\mu_{BAMM}$) as response variables. The numerical values for the model estimates using the MCC tree and HPD intervals of estimates from 100 random trees can be found in the ESM. Density curves are based on estimates from 100 trees and the circle below with error bars is the estimate and 95 % CIs from the MCC tree. For this figure we removed outliers from estimates coming from the 100 random trees for BAMM models in order to interpret the MCC 95 % CIs. \label{model_results}](Figures/ModelEstimates.pdf)
